Hydra Battles and Cie

### 2020-11-15 doc/hydras.tex
restructuration of latex files

\author{Pierre Castéran\\ LaBRI, Univ. Bordeaux, CNRS UMR 5800, Bordeaux-INP\\with contributions by \'Evelyne Contejean,  Florian Hatat and Pascal Manoury}
\date{\today}
\title{Hydra Battles in Coq}
\makeindex
\begin{document}
\maketitle

@@

Among all the theorems proved with the help of proof assistants like \coq{}, \isabelle{}, \hol{}, etc.,
several statements and proofs  share some interesting features:
\begin{itemize}
\item Their statements are easy to understand, even by non-mathematicians
\item Their proof requires some non-trivial mathematical tools
\item Their mechanization on computer presents some methodological interest.
\end{itemize}

@@

Hydra games (a.k.a. \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians:
L. Kirby and J. Paris~\cite{KP82}: \emph{Accessible Independence Results for Peano Arithmetic}. 
Although the mathematical contents of this 
paper are quite advanced, the rules of hydra battles are very easy to understand. There are now several sites on Internet where you can find tutorials on hydra games, together with simulators you can play with. See, for instance, the page written by Andrej Bauer~\cite{bauer2008}.

@@

Here, we present a development  for the \coq{} proof assistant, after the work of Kirby and Paris. This formalization contains the following main parts:

\begin{itemize}
\item Representation in \coq{} of hydras and hydra battles
\item A proof that every battle is finite and won by Hercules. This proof is based on a \emph{variant} which maps any hydra to an ordinal strictly less than $\epsilon_0$ and is strictly decreasing along any battle.

\item Using a combinatorial toolkit designed by J.~Ketonen and R.~Solovay~\cite{KS81}, we prove that, for any ordinal $\mu<\epsilon_0$, there exists no such variant mapping any hydra to an ordinal stricly less than $\mu$. Thus, the complexity of $\epsilon_0$ is really needed in the previous proof.

\item We prove a relation between the length of a ``classic''  class of  battles \footnote{This class is also called \emph{standard} in this document (text and proofs). The \emph{replication factor} of the hydra is exactly $i$ at the $i$-th round of the battle (see Sect~\vref{sect:replication-def}).}
and the Wainer-Hardy hierarchy of ``rapidly growing functions'' $H_\alpha$~\cite{Wainer1970}. The considered class of battles, which we call \emph{standard}  is the most considered one in the scientific  litterature(including popularization).
\end{itemize}

Simply put, this document tries to combines the scientific interest of two articles~\cite{KP82, KS81} and a book~\cite{schutte} with the playful activity of truly proving theorems.
We hope  that such a work, besides exploring a nice piece of discrete maths, 
will show how \coq{} and its standard library are well fitted to help us to understand some non-trivial mathematical developments, and also to experiment the constructive parts of  the proof through functional programming.

 We also hope to provide a little clarification on infinity (both potential and actual) through the notions of function, computation, limit,
 types and proofs.

\section{Remarks}

\subsection{Difference from Kirby and Paris's Work}
In~\cite{KP82}, Kirby and Paris showed  that there is no proof of termination of all hydra battles in Peano Arithmetic (PA).
Since we are used to writing proofs in higher order logic, the restriction to PA was quite unnatural for us. So we chosed to prove another statement without any reference to PA, by considering a class of proofs indexed by ordinal numbers upto $\epsilon_0$.

\subsection{Trust in our Proofs?}
\label{sect:trust-in-proofs}

@@

\subsection{Main References}

In our development, we adapt the definitions and prove many theorems which
we found in the following articles. 
\begin{itemize}
\item ``Accessible independence results for Peano arithmetic''  by Laurie Kirby and Jeff Paris~\cite{KP82}
\item ''Rapidly growing Ramsey Functions'' by Jussi Ketonen and Robert Solovay~\cite{KS81}
\item ``The Termite and the Tower'', by Will Sladek~\cite{Sladek07thetermite}
\item Chapter V of ``Proof Theory'' by Kurt Schütte~\cite{schutte}
\end{itemize}

@@

\subsection{State of the Development}
The \coq{} scripts herein are in constant development since our contribution~\cite{CantorContrib} on  notations for the ordinals $\epsilon_0$ and $\Gamma_0$.
We added new material : axiomatic definition of countable ordinals after Schütte~\cite{schutte}, combinatorial aspects of $\epsilon_0$, after Ketonen and Solovay~\cite{KS81} and Kirby and Paris~\cite{KP82}, recent \coq{} technology: type classes, equations, etc.

We are now working in order to make clumsy proofs more readable, simplify definitions, and ``factorize'' proofs as much as possible. 
Many possible improvements are suggested as ``todo''s or ``projects'' in this text.

\subsection{Contributions}
Many thanks to \'Evelyne Contejean,  Florian Hatat and Pascal Manoury  for their contribution. \'Evelyne contributed libraries on the recursive path ordering (\emph{rpo}) for proving the well-foundedness of our representation of $\epsilon_0$ and $\Gamma_0$.
Florian Hatat proved many useful lemmas on countable sets, which we used in our adaptation of Schütte's formalization of countable ordinals. Pascal Manoury is integrating the ordinal $\omega^\omega$ into our hierarchy of ordinal notations.

@@

\section{Acknowledgements}
\label{sec:orgheadline5}
    Many thanks to David Ilcinkas, Sylvain Salvati, Alan Schmitt and Théo Zimmerman for their help on the elaboration of this document, and to the
 members of the \emph{Formal Methods} team at laBRI for their helpful comments 
on an oral presentation of this work.

@@

\section{How to Install the Libraries}
\label{sec:orgheadline4}
\begin{itemize}
\item The present distribution has been checked with version 8.11.0 of the Coq proof assistant, with the plug-ins \texttt{coq-paramcoq} and \texttt{coq-equations}.

\item Just go into the root directory, and type "\texttt{make}"
\end{itemize}

@@

\section{Future Work (projects)}
\index{Projects}

This document and the proof scripts are far from being complete.

First, there must be a lot of typos to correct, references and index items to add. Many proofs are too complex and should be simplified, etc.

The following extensions are planned, but help is needed:

\begin{itemize}
\item Semi automatic tactics for proving inequalities $\alpha < \beta$, even when $\alpha$ and $\beta$ are not closed terms.
\item Extension to $\Gamma_0$ (in Veblen normal form)
\item More lemmas about hierarchies of rapidly growing functions, and their relationship 
    with primitive recursive functions and provability in Peano arithmetic 
(following~\cite{KS81, KP82}).
\item From \coq's point of view, this development could be used as an illustration of the evolution of the software, every time new libraries and sets of tactics could help to simplify the proofs.
\end{itemize}

%-----------------------------------------------------------

\chapter{Hydras and Hydra Games}
\label{sec:orgheadline91}

This chapter is dedicated to the representation of hydras and rules of the hydra game in \coq's specification language:~\gallina.

Technically, a \emph{hydra} is just a finite ordered tree, each node of which 
has any number of sons. Note that, contrary to the computer science tradition, we will show the hydras 
with the heads up and the foot (i.e., the root of the tree) down.
Fig.~\ref{fig:Hy} represents such  a hydra, which will be referred to as \texttt{Hy} in our examples (please look at the 
module~\href{../theories/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}).

\begin{figure}[h]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4) {$\bullet$};
\node (N3) at (2,6) {$\bullet$};
\node (H0) at (0,2) {$\Smiley[2][vertfluo]$};
\node (H1) at (0,8) {$\Smiley[2][vertfluo]$};
\node (H2) at (4,8) {$\Smiley[2][vertfluo]$};
\node (H4) at (4,2) {$\Smiley[2][vertfluo]$};
\node (H5) at (4,4) {$\Smiley[2][vertfluo]$};
\draw (foot) -- (N1)[very thick] ;
\draw (N1) -- (N2);
\draw (N2) -- (N3);
\draw (N3) to [bend left= 10]  (H1) ;
\draw (N3) to [bend right= 16] (H2);
\draw (foot) to [bend left= 10]  (H0) ;
\draw (foot) to [bend right = 10] (H4) ;
\draw (N1) to [bend right= 16] (H5);
\end{tikzpicture}
\caption{The hydra Hy \label{fig:Hy}}
\end{figure}

We use a specific vocabulary for talking about hydras. Table~\ref{tab:hyd2tree} shows the correspondance between our terminology and the usual vocabulary for trees in computer science.

\begin{figure}[h]
  \centering
  \begin{tabular}{ll}
Hydras & Finite rooted trees\\
\hline
foot & root\\
head & leaf\\
node & node\\
segment  & (directed) edge \\
sub-hydra & subtree\\
daughter & immediate subtree\\
\end{tabular}
  \caption{Translation from hydras to trees}
  \label{tab:hyd2tree}
\end{figure}

The hydra \texttt{Hy} has a \emph{foot} (below), five \emph{heads}, and eight \emph{segments}. 
We leave it to the reader to define various parameters such as the height, the size, the highest arity (number of sons of a node) of a hydra. In our example, these parameters have the respective values : $4$, $9$ and $3$.

\subsection{The Rules of the Game}

\label{sec:orgheadline44}
\label{sect:replication-def}

A \emph{hydra battle} is a fight between Hercules and the Hydra. 
More formally, a  battle is a sequence of \emph{rounds}.
At each round:
\begin{itemize}
\item If the hydra is composed of just one head, the battle is finished
and  Hercules is the winner.
\item Otherwise, Hercules chops off \emph{one} head of the hydra,

\begin{itemize}
\item If the head is at distance 1 from the foot, the head is just lost by the hydra, with no more reaction.
\item Otherwise, let us denote by \(r\) the node that was at distance \(2\) from 
the removed head in the direction of the foot,  and consider the  sub-hydra \(h'\) of \(h\), whose  root is \(r\) \footnote{$h'$ will be called ``the wounded part of the hydra'' in the subsequent text. In Figures~\vref{fig:Hy2} and ~\vref{fig:Hy4}, this sub-hydra  is displayed in red.}. Let $n$ be some natural number.
Then $h'$ is replaced by  $n+1$ of copies of \(h'\) which share the same root $r$.
 The \emph{replication factor} $n$ may be different (and generally is)   at each round of the fight.
It may be chosen by the hydra, according to its strategy, or imposed by some 
particular rule. In many presentations of hydra battles, this number is increased by $1$ at each round. In the following presentation, we will also consider battles where the hydra is free to choose its ~replication factor at each round of the battle\footnote{Let us recall that, if the chopped-off head was at distance 1 from the foot, the replication factor is meaningless.}.
\end{itemize}
\end{itemize}

Note that the description given in~\cite{KP82} of the replication process in hydra battles is also  semi-formal.

\label{original-rules}

\begin{quote}
  ``From the node that used to be attached to the head which was just chopped off, traverse one 
segment towards the root until the next node is reached. From this node sprout $n$ replicas of 
that part of the hydra (after decapitation) which is ``above'' the segment just traversed, i.e., those 
nodes and segments from which, in order to reach the root, this segment would have to be 
traversed. If the head just chopped off had the root of its nodes, no new head is grown. ''
\end{quote}

Moreover, we note that this description is in \emph{imperative} terms. In order to build a formal  study of the properties of hydra battles, we prefer to use a mathematical vocabulary, i.e., graphs, relations, functions, etc.
Thus, the replication process will be represented as a binary relation on a data type \texttt{Hydra},
linking the state of the hydra \emph{before} and \emph{after} the transformation.
A battle will thus be represented as a sequence of terms of type \texttt{Hydra}, respecting the rules of the game.

\subsection{Example}
Let us start a battle between Hercules and the hydra \texttt{Hy} of Fig.~\ref{fig:Hy}.

At the first round, Hercules choses to chop off the rightmost head of \texttt{Hy}.
Since this head is near the floor, the hydra loses this head. Let us call 
 \texttt{Hy'} the resulting state of the hydra, represented in Fig.~\vref{fig:Hy-prime}.

Next, assume Hercules choses to chop off one of the two highest heads of \texttt{Hy'}, for instance the rightmost one. Fig.~\vref{fig:Hy2} represents the rotten neck in dashed lines, and the part that will be replicated in red. Assume also that the hydra decides to add 4 copies of the red part\footnote{In other words, the replication factor at this round is equal to $4$.}. We obtain a new state \texttt{Hy''} depicted in Fig.~\ref{fig:Hy3}.

\begin{figure}[h]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4) {$\bullet$};
\node (N3) at (2,6) {$\bullet$};
\node (H0) at (0,2) {$\Smiley[2][vertfluo]$};
\node (H1) at (0,8) {$\Smiley[2][vertfluo]$};
\node (H2) at (4,8) {$\Smiley[2][vertfluo]$};
\node (H5) at (4,4) {$\Smiley[2][vertfluo]$};
%\node (H4) at (6,0) {$\Xey[2][lightgray]$};
\draw (foot) -- (N1)[very thick] ;
\draw (N1) -- (N2);
\draw (N2) -- (N3) ;
\draw (N3) to [bend left= 10]  (H1) ;
\draw (N3) to [bend right= 16] (H2);
\draw (foot) to [bend left= 10]  (H0) ;
\draw (N1) to [bend right= 16] (H5);
\end{tikzpicture}

\caption{Hy': the state  of Hy after one round \label{fig:Hy-prime}}
\end{figure}

\begin{figure}[hp]
\centering
\begin{tikzpicture}[very thick, scale=0.5]

\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4)  {{\color{lightred}$\bullet$}};
\node (N3) at (2,6) {{\color{lightred}$\bullet$}};
\node (H0) at (0,2) {$\Smiley[2][vertfluo]$};
\node (H1) at (0,8) {$\Sey[2][lightred]$};
%\node (H2) at (5,0) {$\Xey[2][lightgray]$};
\node (H5) at (4,4) {$\Smiley[2][vertfluo]$};
\node (ex) at (5,8) {};
\draw (foot) -- (N1)[very thick] ;
\draw (N1) -- (N2);
\draw  (N2) -- (N3)[draw=lightred];
\draw (N3) to   [bend left= 10](H1) [draw=lightred];
\draw [dashed] (N3) to [bend left= 10](ex);
\draw (foot) to [bend left= 10]  (H0) ;
\draw (N1) to [bend right= 16] (H5);
\end{tikzpicture}
\caption{A second beheading}
\label{fig:Hy2}
\end{figure}

\begin{figure}[hp]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4) {$\bullet$};
\node (N3) at (2,6) {{\color{lightred}$\bullet$}};
\node (H1) at (0,8) {$\Smiley[2][vertfluo]$};
\node (H11) at (2,8) {$\Smiley[2][vertfluo]$};
\node (H12) at (4,8) {$\Smiley[2][vertfluo]$};
\node (H13) at (6,8) {$\Smiley[2][vertfluo]$};
\node (H14) at (8,8) {$\Smiley[2][vertfluo]$};

\node (N3) at (1,6) {$\bullet$};
\node (N31) at (2,6) {$\bullet$};
\node (N32) at (3,6) {$\bullet$};
\node (N33) at (4,6) {$\bullet$};
\node (N34) at (5,6) {$\bullet$};

\node (H0) at (0,2) {$\Smiley[2][vertfluo]$};
\node (H5) at (4,4) {$\Smiley[2][vertfluo]$};
\draw (foot) -- (N1)[very thick] ;
\draw (N1) -- (N2);
\draw (N2) -- (N3);
\draw (N2) -- (N31);
\draw (N2) -- (N32);
\draw (N2) -- (N33);
\draw (N2) -- (N34);
\draw (N3) to   [bend left= 10](H1) ;
\draw (N31) to   [bend left= 10](H11) ;
\draw (N32) to   [bend left= 10](H12) ;
\draw (N33) to   [bend left= 10](H13) ;
\draw (N34) to   [bend left= 10](H14) ;
\draw (foot) to [bend left= 10]  (H0) ;
\draw (N1) to [bend left= 10]  (H5) ;
\end{tikzpicture}
\caption{Hy'', the state of Hy after two rounds \label{fig:Hy3}}
\end{figure}

Figs.~\ref{fig:Hy4} and~\vref{fig:Hy5} represent a possible third round of the battle, with a replication factor equal to $2$. Let us call \texttt{Hy'''} the state of the hydra after that third round.

\begin{figure}[hp]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

\node (foot) at (2,0)  {{\color{lightred}$\bullet$}};
\node (N1) at (2,2) {{\color{lightred}$\bullet$}};
\node (N2) at (2,4) {{\color{lightred}$\bullet$}};
\node (N3) at (2,6) {{\color{lightred}$\bullet$}};
\node (exN4) at (4,4) {};
\node (H1) at (0,8) {$\Sey[2][lightred]$};
\node (H11) at (2,8) {$\Sey[2][lightred]$};
\node (H12) at (4,8) {$\Sey[2][lightred]$};
\node (H13) at (6,8) {$\Sey[2][lightred]$};
\node (H14) at (8,8) {$\Sey[2][lightred]$};

\node (N3) at (1,6) {{\color{lightred}$\bullet$}};
\node (N31) at (2,6) {{\color{lightred}$\bullet$}};
\node (N32) at (3,6) {{\color{lightred}$\bullet$}};
\node (N33) at (4,6) {{\color{lightred}$\bullet$}};
\node (N34) at (5,6) {{\color{lightred}$\bullet$}};

\node (H0) at (0,2) {$\Smiley[2][vertfluo]$};
%\node (H5) at (4,0) {$\Xey[2][lightgray]$};
\draw (foot) -- (N1)[very thick,draw=lightred] ;
\draw (N1) -- (N2)[draw=lightred];
\draw (N2) -- (N3)[draw=lightred];
\draw (N2) -- (N31)[draw=lightred];
\draw (N2) -- (N32)[draw=lightred];
\draw (N2) -- (N33)[draw=lightred];
\draw (N2) -- (N34)[draw=lightred];
\draw (N3) to   [bend left= 10](H1) [draw=lightred];
\draw (N31) to   [bend left= 10](H11) [draw=lightred];
\draw (N32) to   [bend left= 10](H12) [draw=lightred];
\draw (N33) to   [bend left= 10](H13) [draw=lightred];
\draw (N34) to   [bend left= 10](H14) [draw=lightred];
\draw (foot) to [bend left= 10]  (H0) ;
\draw [dashed] (N1) to  [bend left= 10](exN4);
\end{tikzpicture}
\caption{A third beheading (wounded part in red) \label{fig:Hy4}}
\end{figure}

\begin{figure}[hp]
\centering
\begin{tikzpicture}[very thick, scale=0.4]

\node (foot) at (10,0) {$\bullet$};

\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4) {$\bullet$};
\node (N3) at (2,6) {{\color{lightred}$\bullet$}};
\node (H1) at (0,8) {$\Smiley[1][vertfluo]$};
\node (H11) at (2,8) {$\Smiley[1][vertfluo]$};
\node (H12) at (4,8) {$\Smiley[1][vertfluo]$};
\node (H13) at (6,8) {$\Smiley[1][vertfluo]$};
\node (H14) at (8,8) {$\Smiley[1][vertfluo]$};

\node (N3) at (1,6) {$\bullet$};
\node (N31) at (2,6) {$\bullet$};
\node (N32) at (3,6) {$\bullet$};
\node (N33) at (4,6) {$\bullet$};
\node (N34) at (5,6) {$\bullet$};

\node (H0) at (-3,3) {$\Smiley[1][vertfluo]$};

\draw (foot) to [bend left=10] (N1)[very thick] ;
\draw (N1) -- (N2);
\draw (N2) -- (N3);
\draw (N2) -- (N31);
\draw (N2) -- (N32);
\draw (N2) -- (N33);
\draw (N2) -- (N34);
\draw (N3) to   [bend left= 10](H1) ;
\draw (N31) to   [bend left= 10](H11) ;
\draw (N32) to   [bend left= 10](H12) ;
\draw (N33) to   [bend left= 10](H13) ;
\draw (N34) to   [bend left= 10](H14) ;
\draw (foot) to [bend left = 15]  (H0) ;

% second copy 
\node (N01) at (12,2) {$\bullet$};
\node (N02) at (12,4) {$\bullet$};
\node (N03) at (12,6) {{\color{lightred}$\bullet$}};
\node (H001) at (10,8) {$\Smiley[1][vertfluo]$};
\node (H0011) at (12,8) {$\Smiley[1][vertfluo]$};
\node (H0012) at (14,8) {$\Smiley[1][vertfluo]$};
\node (H0013) at (16,8) {$\Smiley[1][vertfluo]$};
\node (H0014) at (18,8) {$\Smiley[1][vertfluo]$};

\node (N03) at (11,6) {$\bullet$};
\node (N031) at (12,6) {$\bullet$};
\node (N032) at (13,6) {$\bullet$};
\node (N033) at (14,6) {$\bullet$};
\node (N034) at (15,6) {$\bullet$};

\draw (foot) -- (N01)[very thick] ;
\draw (N01) -- (N02);
\draw (N02) -- (N03);
\draw (N02) -- (N031);
\draw (N02) -- (N032);
\draw (N02) -- (N033);
\draw (N02) -- (N034);
\draw (N03) to   [bend left= 10](H001) ;
\draw (N031) to   [bend left= 10](H0011) ;
\draw (N032) to   [bend left= 10](H0012) ;
\draw (N033) to   [bend left= 10](H0013) ;
\draw (N034) to   [bend left= 10](H0014) ;

% third copy 
\node (N001) at (22,2) {$\bullet$};
\node (N002) at (22,4) {$\bullet$};
\node (N003) at (22,6) {{\color{lightred}$\bullet$}};
\node (H001) at (20,8) {$\Smiley[1][vertfluo]$};
\node (H0011) at (22,8) {$\Smiley[1][vertfluo]$};
\node (H0012) at (24,8) {$\Smiley[1][vertfluo]$};
\node (H0013) at (26,8) {$\Smiley[1][vertfluo]$};
\node (H0014) at (28,8) {$\Smiley[1][vertfluo]$};

\node (N003) at (21,6) {$\bullet$};
\node (N0031) at (22,6) {$\bullet$};
\node (N0032) at (23,6) {$\bullet$};
\node (N0033) at (24,6) {$\bullet$};
\node (N0034) at (25,6) {$\bullet$};

\draw (foot) -- (N001)[very thick] ;
\draw (N001) -- (N002);
\draw (N002) -- (N003);
\draw (N002) -- (N0031);
\draw (N002) -- (N0032);
\draw (N002) -- (N0033);
\draw (N002) -- (N0034);
\draw (N003) to   [bend left= 10](H001) ;
\draw (N0031) to   [bend left= 10](H0011) ;
\draw (N0032) to   [bend left= 10](H0012) ;
\draw (N0033) to   [bend left= 10](H0013) ;
\draw (N0034) to   [bend left= 10](H0014) ;
\end{tikzpicture}
\caption{The configuration Hy''' of Hy \label{fig:Hy5}}
\end{figure}
\FloatBarrier

We leave it to the reader  to guess the following  rounds of the battle \dots
 % Please keep in mind that, in this 
% the hydra is free to chose any number of replications at each time, whereas
% Hercules chops only one head per round.

% Let us precise that, in this game, Hercules wins if the hydra is eventually reduced 
% to a single head. 
% We know from~\cite{KP82} that, whichever the initial configuration of the
% hydra, and the strategies of both players, Hercules eventually wins. The 
% aforementionned paper shows also that there do not exist any \emph{simple} proof of this result.

\section{Hydras and their Representation in \emph{Coq}}
\label{sec:orgheadline48}

In order to describe trees where each node can have an arbitrary (but finite) number of sons, it usual to define a type where each node carries a \emph{forest}, \emph{i.e} a list of trees
(see for instance Chapter 14, pages 400-406 of \cite{BC04}).

For this purpose, we define two mutual \emph{ad-hoc}  inductive types, where \texttt{Hydra} is the main type, and \texttt{Hydrae} is a helper for describing finite sequences of hydra.
\label{types:Hydra}
\label{types:Hydrae}

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#Hydra}{Hydra.Hydra\_Definitions}}

\begin{Coqsrc}
Inductive Hydra : Set :=
| node :  Hydrae -> Hydra
with Hydrae : Set :=
| hnil : Hydrae
| hcons : Hydra -> Hydrae -> Hydrae.
\end{Coqsrc}

%\index{To do}
\index{Projects}

\begin{project}
Look for an existing library on trees with nodes of arbitrary arity, in order to replace  this ad-hoc type with something more generic.
\end{project}

\index{Projects}

\begin{project}

 Another very similar representation could use the \texttt{list} type family instead of the specific 
type \texttt{Hydrae}:

\begin{Coqalt}
Module Alt.

Inductive Hydra: Set :=
  hnode (daughters : list Hydra).

End Alt.
\end{Coqalt}

Using this representation, re-define all the constructions of this chapter.
You will probably have to use patterns described for instance in~\cite{BC04} or the archives of the Coq-club~\cite{Coq}.

\end{project}

\index{Projects}

\begin{project}
The type \texttt{Hydra} above describes hydras as \emph{plane trees}, i.e., as drawn on a sheet of paper or computer screen. Thus, hydras are \emph{oriented},
and it is appropriate to consider a \emph{leftmost} or \emph{rightmost} head of
the beast. It could be interesting to consider another representation, in which
every non-leaf node has a \emph{multi-set} -- not an ordered list -- of daughters.
\end{project}

\subsubsection{Abbreviations}

We provide several notations for hydra patterns  which occur often in our developments.

\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#head}{Hydra.Hydra\_Definitions}}

\begin{Coqsrc}
(** heads *)
Notation head := (node hnil).

(** nodes  with 1, 2 or 3 daughters *)
Notation hyd1 h := (node (hcons h hnil)).
Notation hyd2 h h' := (node (hcons h (hcons h' hnil))).
Notation hyd3 h h' h'' := 
                   (node (hcons h (hcons h' (hcons h'' hnil)))).
\end{Coqsrc}

For instance, the hydra \texttt{Hy}  of Figure~\vref{fig:Hy} is defined in \emph{Gallina} as follows:

\vspace{4mm}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Examples.html\#Hy}{Hydra.Hydra\_Examples}}

\begin{Coqsrc}
Example Hy := hyd3 head
                   (hyd2
                      (hyd1 
                         (hyd2 head head))
                      head) 
                   head.
\end{Coqsrc}

Hydras quite frequently contain  multiple  copies of the same pattern. The following functions
will help us to describe and reason about replications in hydra battles.

\vspace{4pt}

\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#hcons_mult}{Hydra.Hydra\_Definitions}}

\begin{Coqsrc}
Fixpoint hcons_mult (h:Hydra)(n:nat)(s:Hydrae):Hydrae :=
  match n with 
  | O => s
  | S p => hcons h (hcons_mult h p s)
  end.

(** hydra with n copies of the same daughter *)

Definition hyd_mult h n :=
  node (hcons_mult h n hnil).
\end{Coqsrc}

\vspace{4mm}

For instance, the hydra $Hy''$ of Fig~\vref{fig:Hy3}  can be defined in \coq{} as follows:

\vspace{4pt}

\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}}

\begin{Coqsrc}
Example Hy'' := 
     hyd2 head
          (hyd2 (hyd_mult (hyd1 head) 5)
                head).
\end{Coqsrc}

\subsubsection{Recursive Functions on type \texttt{Hydra}}
\label{sec:orgheadline41}
\label{sec:hsize-def}

When defining a recursive function over the type \texttt{Hydra}, one has to consider the three constructors 
\texttt{node}, \texttt{hnil} and \texttt{hcons} of the mutually inductive types \texttt{Hydra} and \texttt{Hydrae}. 
Let us define for instance the function which  computes the number of nodes of any hydra:

\vspace{4pt}
\noindent
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}

\begin{Coqsrc}
Fixpoint hsize (h:Hydra) : nat :=
  match h with node l => S (lhsize l)
  end
with lhsize l : nat :=
  match l with hnil => 0
            | hcons h hs => hsize h + lhsize hs 
  end.

 Compute hsize Hy.
\end{Coqsrc}

\begin{Coqanswer}
 = 9
     : nat 
\end{Coqanswer}

Likewise, the \emph{height} (maximum distance between the foot and a head) 
is defined by mutual recursion:

\begin{Coqsrc}
Fixpoint height  (h:Hydra) : nat :=
  match h with node l => lheight l
  end
with lheight l : nat :=
  match l with 
  | hnil => 0
  | hcons h hs => Max.max (S (height h)) (lheight hs)
  end.
\end{Coqsrc}

\begin{Coqsrc}
Compute height Hy.
\end{Coqsrc}

\begin{Coqanswer}
 = 4
     : nat  
\end{Coqanswer}

\index{Exercises}

\begin{exercise}
Define a function \texttt{max\_degree: Hydra $\arrow$ nat} which  returns the highest degree of a node in any hydra. For instance, the evaluation of the term \texttt{(max\_degree Hy)} should return $3$.
\end{exercise}

\subsection{Induction Principles for Hydras}
\label{sec:orgheadline42}

In this section, we show how induction principles are used to prove properties on the type 
\texttt{Hydra}. Let us consider for instance the following statement:
\begin{quote}
  `` The height of any hydra is strictly less than its size. ''
\end{quote}

\subsubsection{A failed Attempt}

One may try to use the default tactic of proof by induction, which corresponds to an application of the automatically  generated  induction principle for  type \texttt{Hydra}:

\begin{Coqanswer}
Hydra_ind :
forall P : Hydra -> Prop,
(forall h : Hydrae, P (node h)) -> forall h : Hydra, P h
\end{Coqanswer}

Ler us start a simple proof by induction.

\vspace{4pt}
\noindent
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}}

\begin{Coqbad}
Module Bad.

Lemma height_lt_size (h:Hydra) :
  height h <= hsize h.
Proof.
  induction h as [s].
\end{Coqbad}

\begin{Coqanswer}
1 subgoal, subgoal 1 (ID 11)

  s : Hydrae
  ============================
   height (node s) <= hsize (node s)
\end{Coqanswer}

We might be tempted to do an induction on the sequence \texttt{s}:

% \begin{Coqbad}
% induction  s as [ | h s'].
%   -   cbn; auto with arith.
%   - 
% \end{Coqbad}

\begin{Coqanswer}
 1 focused subgoal
(unfocused: 0), subgoal 1 (ID 19)

  h : Hydra
  s' : Hydrae
  IHs' : height (node s') <= hsize (node s')
  ============================
   height (node (hcons h s')) <= hsize (node (hcons h s'))

Note that the displayed subgoal does not contain any assumption on \texttt{h}, thus there is no way to 
infer any property about the height and size of the hydra \texttt{(hcons h t)}.

\begin{Coqbad}
Abort.

\subsubsection{A Principle of Mutual Induction}
In order to get an appropriate induction scheme for the types 
\texttt{Hydra} and \texttt{Hydrae}, we can use  \coq{}'s  command \texttt{Scheme}.

\index{Coq!Techniques!Mutually inductive types}
\index{Coq!Commands!Scheme}

\begin{Coqsrc}
Scheme Hydra_rect2 := Induction for Hydra Sort Type
with Hydrae_rect2 := Induction for Hydrae Sort Type.
\end{Coqsrc}

\begin{Coqsrc}
Check Hydra_rect2.
\end{Coqsrc}

\begin{Coqanswer}
Hydra_rect2
 : forall (P : Hydra -> Type) (P0 : Hydrae -> Type),
   (forall h : Hydrae, P0 h -> P (node h)) ->
    P0 hnil ->
    (forall h : Hydra, P h -> 
             forall h0 : Hydrae, P0 h0 -> P0 (hcons h h0)) ->
    forall h : Hydra, P h
\end{Coqanswer}

\subsubsection{A Correct Proof}

>>>>

\author{Pierre Castéran\\ LaBRI, Univ. Bordeaux, CNRS UMR 5800, Bordeaux-INP\\with contributions by Yves Bertot, \'Evelyne Contejean,  Florian Hatat and Pascal Manoury}
\date{\today}
\title{Hydra Battles and Cie}
\makeindex
\begin{document}
\maketitle

@@

Among all the theorems proved with the help of proof assistants like \coq{}, \isabelle{}~\cite{isabelle}, \hol{}~\cite{HOL}, etc.,
several statements and proofs  share some interesting features:
\begin{itemize}
\item Their statements are easy to understand, even by non-mathematicians
\item Their proof requires some non-trivial mathematical tools
\item Their mechanization on computer presents some methodological interest.
\end{itemize}

@@

In this document, we present two examples which seem to have the above properties.

\begin{itemize}
\item Hydra games (a.k.a. \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians:
L. Kirby and J. Paris~\cite{KP82}: \emph{Accessible Independence Results for Peano Arithmetic}. 
Although the mathematical contents of this 
paper are quite advanced, the rules of hydra battles are very easy to understand. There are now several sites on Internet where you can find tutorials on hydra games, together with simulators you can play with. See, for instance, the page written by Andrej Bauer~\cite{bauer2008}.

@@

\item In the second part, we are interested in computing $x^n$ with the least number of multiplications as possible. We use the notion of \emph{addition chains}~\cite{brauer1939,DBLP:journals/ipl/BerstelB87}, to generate efficient certified exponentiation functions.
\end{itemize}

%%%%% ICI

\subsection{Trust in our Proofs}
\label{sect:trust-in-proofs}

@@

@@

\subsection{Contributions}
Many thanks to Yves Bertot, 
\'Evelyne Contejean,  Florian Hatat and Pascal Manoury  for their contribution. 
Yves made nice optimizations  to algorithms presented in Chapter~\ref{chapter-powers}.
\'Evelyne contributed libraries on the recursive path ordering (\emph{rpo}) for proving the well-foundedness of our representation of $\epsilon_0$ and $\Gamma_0$.
Florian Hatat proved many useful lemmas on countable sets, which we used in our adaptation of Schütte's formalization of countable ordinals. Pascal Manoury is integrating the ordinal $\omega^\omega$ into our hierarchy of ordinal notations.

@@

\section{Acknowledgements}
\label{sec:orgheadline5}
    Many thanks to Yves Bertot, David Ilcinkas, Sylvain Salvati, Alan Schmitt and Théo Zimmerman for their help on the elaboration of this document, and to the
 members of the \emph{Formal Methods} team at laBRI for their helpful comments 
on an oral presentation of this work.

@@

\section{How to Install the Libraries}
\label{sec:orgheadline4}
\begin{itemize}
\item The present distribution has been checked with version 8.12.0 of the Coq proof assistant, with the plug-ins \texttt{coq-paramcoq}, \texttt{coq-equations}
and \texttt{coq-mathcomp-algebra}.

\item Just go into the root directory, and type "\texttt{make}" ("\texttt{make -k}"  in case of an incomplete installation).
\end{itemize}

@@

\part{Hydras and Ordinals}

\include{part-hydras}

\part{Some Funny Algorithms}

\include{chapter-powers}

\part{Appendices}
